---
title: "Lab 02 - Plastic waste"
author: "Cailey Fay "
date: "10.6.25"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

Warm up: Top left pane is for scripts / syntax / markdown files. Bottom left is the console where you can punch in commands and R spits stuff out. Top right pane is the environment, lets you see what data files are in operation, and under the git tab you can make your way to committing changes. Bottom right pane is where you can see plots, info on how to use a particular function or package, and for accessing files. 

There are 240 observations in the plastic waste dataset. 

### Exercise 1

To get the distribution of plastic waste per capita: 

```{r plastic-waste-continent}
#plain old histogram of plastic waste per capita 
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)

```

To see the countries where it is > 3.5 kg/person:
```{r}
plastic_waste%>%
  filter(plastic_waste_per_cap > 3.5) 

#creating the df that filters out the outliers 

df_filtered <- plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5)
```
1.1 Plotting the distribution of plastic waste per capita faceted by continent. 
```{r}
  ggplot(data = df_filtered, 
         aes(x=plastic_waste_per_cap)) +
 geom_histogram(binwidth = .1) +
facet_wrap(~continent)

```
Conclusion: African and Asian counties are positively skewed, and tend to have lower plastic waste per cap than North America and Europe. Less can be said about Oceania and South America, since there are low frequencies / not a lot of countries with data points. 

With density plots 
```{r}
ggplot(data = df_filtered, 
       aes(x=plastic_waste_per_cap)) + 
  geom_density()

#getting fancier 
ggplot(data=df_filtered, 
       mapping = aes(x=plastic_waste_per_cap, 
                     color = continent)) + 
  geom_density()

#even fancier 
ggplot(data=df_filtered,
       mapping = aes(x=plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) + geom_density()

#Better but not quite there yet 
ggplot(data=df_filtered, 
       mapping = aes(
         x=plastic_waste_per_cap, 
         color = continent, 
         fill = continent)) + 
  geom_density(alpha=0.7)
```

### Exercise 2
2.1 Recreating the plots with lower alpha 
```{r plastic-waste-density}
ggplot(data=df_filtered,
       mapping = aes(
         x=plastic_waste_per_cap, 
         color = continent, fill = continent)) + 
  geom_density(alpha=.2)
```
Nicer to look at. 

2.2 Color and fill are aesthetics because 

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.

```{r plastic-waste-violin}
ggplot(
  data = df_filtered,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_boxplot()

geom_violin(
  mapping = NULL,
  data = NULL,
  stat = "ydensity",
  position = "dodge",
  trim = TRUE,
  bounds = c(-Inf, Inf),
  quantile.colour = NULL,
  quantile.color = NULL,
  quantile.linetype = 0L,
  quantile.linewidth = NULL,
  scale = "area",
  na.rm = FALSE,
  orientation = NA,
  show.legend = NA,
  inherit.aes = TRUE
)

stat_ydensity(
  mapping = NULL,
  data = NULL,
  geom = "violin",
  position = "dodge",
  orientation = NA,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  trim = TRUE,
  scale = "area",
  drop = TRUE,
  bounds = c(-Inf, Inf),
  quantiles = c(0.25, 0.5, 0.75),
  na.rm = FALSE,
  show.legend = NA,
  inherit.aes = TRUE
)

#confused by this 
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r plastic-waste-mismanaged}
ggplot(data=df_filtered, mapping= aes(x=plastic_waste_per_cap, y=mismanaged_plastic_waste_per_cap) ) + geom_point()
```

```{r plastic-waste-mismanaged-continent}
ggplot(data=df_filtered, mapping= aes(x=plastic_waste_per_cap, y=mismanaged_plastic_waste_per_cap, color=continent) ) + geom_point()
```

```{r plastic-waste-population-total}
ggplot(data=df_filtered, mapping= aes(x=plastic_waste_per_cap, y=total_pop, color=continent) ) + geom_point()

ggplot(
  data = df_filtered,
  mapping = aes(
    x = plastic_waste_per_cap,
    y = total_pop, color=continent
  )
) +
  geom_boxplot()
```

```{r plastic-waste-population-coastal}

```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
CoastalProportion <- df_filtered %>%
  mutate(CP = coastal_pop / total_pop)

ggplot(data = CoastalProportion, mapping = aes(x= CP, y=plastic_waste_per_cap, color=continent)) + geom_point() + geom_smooth() +
  labs(title= "Plastic Waste v Coastal Population Proportion",
       subtitle= "not quite right",
       x="Coastal Proportion",
       y= "Plastic Waste Per Capita")

```


